\chapter{Experimental Protocol and Validation Status}
\label{ch6}

This chapter records the empirical protocol attached to the corrected
SA-VRPTW formulation. A deliberate distinction is preserved
throughout: claims that the current branch \emph{has executed} are
reported with citations to the artefacts in \texttt{data/results/};
claims that depend on the in-flight remote experiment are flagged
explicitly so the manuscript and the software cannot drift out of
alignment.

\section{Setup}

The benchmark covers five Indian cities (Bengaluru, Delhi, Gurugram,
Mumbai, Pune) with depot counts $\{3, 5, 8\}$ scaled to the customer
count $N$ as in Table~\ref{tab:depot-N}. Customer counts are
$N \in \{20, 50, 100, 200\}$. Each (city, $N$, solver) cell is
replicated over multiple random seeds. The MILP is capped at
$N \leq 50$ because CBC does not scale gracefully past that point
inside the 120~s budget.

\begin{table}[ht]
\centering
\caption{Per-instance depot counts as a function of $N$.}
\label{tab:depot-N}
\begin{tabular}{cc}
\toprule
$N$ & depots \\
\midrule
20  & 3 \\
50  & 5 \\
100 & 5 \\
200 & 8 \\
\bottomrule
\end{tabular}
\end{table}

\section{Reproducibility}

The full pipeline (OSM graph $\rightarrow$ BPR congestion $\rightarrow$
calibrated risk $\rightarrow$ Blinkit dark-store clustering
$\rightarrow$ super-graph construction $\rightarrow$ solver
$\rightarrow$ evaluator) is composable through Hydra. A complete
experiment row is reproducible with
\begin{verbatim}
PYTHONPATH=src python scripts/run_experiment.py \
  city=bengaluru solver=ga instance.N=50 \
  instance.seed=42 instance.R_bar=0.40
\end{verbatim}
and the full grid through \texttt{scripts/run\_grid.py}. The grid
runner checkpoints into a manifest at
\texttt{data/results/\_manifest.jsonl} so a remote run can resume
from a crash without recomputing already-finished rows. Every
parameter has a single source-of-truth file under \texttt{conf/}; the
formulation note (\texttt{docs/FORMULATION.md}) and calibration note
(\texttt{docs/BASM\_CALIBRATION.md}) are versioned alongside the
code.

\section{Default-$\varepsilon$ Benchmark}

The default-$\varepsilon$ benchmark sets
$\bar{R} = +\infty$ and $\bar{H} = +\infty$ (no
$\varepsilon$-constraint active) and reports mean $F_1$, mean
wall-clock runtime, and feasibility rate per (city, $N$, solver)
cell. Table~\ref{tab:default-grid} is the planned reporting schema;
empirical entries are populated from \texttt{data/results/} once the
remote run completes.

\begin{table}[ht]
\centering
\caption{Default-$\varepsilon$ benchmark: mean $F_1$ (lower is
better), mean wall-clock seconds, and feasibility rate over the
configured seed set. Empirical values are filled from
\texttt{data/results/} after the iDRAC run.}
\label{tab:default-grid}
\begin{tabular}{lcccc}
\toprule
\multirow{2}{*}{City} & \multicolumn{2}{c}{$N=20$} & \multicolumn{2}{c}{$N=50$} \\
& MILP & GA & MILP & GA \\
\midrule
Bengaluru & --- & --- & --- & --- \\
Delhi     & --- & --- & --- & --- \\
Gurugram  & --- & --- & --- & --- \\
Mumbai    & --- & --- & --- & --- \\
Pune      & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\section{Pareto Sweep: Risk vs.\ $F_1$}

The risk-versus-$F_1$ Pareto frontier is generated by sweeping
$\bar{R}$ at fixed $N = 50$ over
\[
    \bar{R} \in \{0.05, 0.10, 0.20, 0.40, 0.80, 1.60, +\infty\},
\]
producing seven points per city per seed. Non-dominated points are
retained and plotted; dominance filtering is mandatory before any
Pareto figure is published. The operator-facing reading is the
gradient of the frontier: the marginal cost in $F_1$ units of a
one-percentage-point increase in mean per-route survival.

The fleet-wide residential budget $\bar{H}$ is similarly swept over
\[
    \bar{H} \in \{0, 4, 8, 16, 32, +\infty\}
\]
to characterise the cost of network-level vulnerable-street
restriction. The two sweeps may be run jointly to produce a
two-dimensional Pareto surface, although the headline figure in this
thesis is the one-dimensional risk frontier.

\section{Solver Statistical Comparison}

For each (city, $N$) cell, the GA and ALNS incumbents are compared
across paired seeds using the Wilcoxon signed-rank test
\citep{wilcoxon1945} with Bonferroni correction across cells. The
expected pattern from pilot runs is that ALNS dominates GA at
$N \geq 100$ in cities with denser residential penetration, while GA
remains competitive at $N \leq 50$. Statistical claims are populated
from the actual run output rather than asserted in advance.

\section{Behavioural Compliance Sweep}

The behavioural Monte Carlo of Section~\ref{ch5:behav} is run on the
default-$\varepsilon$ Pareto-optimal solutions for each city. The
output is a per-rider survival curve as a function of the compliance
prior parameters $(\alpha, \beta)$; the headline number is the
inter-quartile range across the nine operating points. The
sensitivity sweep is reported as a diagnostic layer in addition to
the analytic per-route survival $e^{-\bar{R}}$.

\section{Validation State of the Branch}

At the time of this manuscript update the refactored code base has
reached a stable software baseline. The smoke test suite is green;
the synthetic end-to-end pipeline is runnable; the BASM calibration
guard rejects uncalibrated configurations at instance build time;
and the GA and ALNS solvers produce strictly feasible incumbents
under the Chapter~3 contract on all tested seeds. What is not yet
claimed in this manuscript is the fully-populated $\varepsilon$-grid
in dominance-filtered form across all five cities; that is the
responsibility of the in-flight remote experiment grid and is
explicitly out of scope for the version of the thesis text
submitted at this checkpoint.

\section{Planned Figure Families}

When the full experiment grid is available, the manuscript will
report eight figure families driven directly by the emitted JSON
rows: (i) solver scalability curves, (ii) $\varepsilon$-swept Pareto
frontiers, (iii) heuristic convergence traces, (iv) crash-survival
box plots, (v) behavioural sensitivity grids, (vi) route-level
risk-versus-$F_1$ scatter views, (vii) residential-edge spatial
overlays, and (viii) normalised cross-city comparisons. Every Pareto
plot is dominance-filtered before display, and every figure is paired
with a committed CSV export to preserve reviewer reproducibility.
