\chapter{Discussion, Limitations, and Conclusion}

\section{Operator-Facing Implications}

The $\varepsilon$-constraint formulation lets a planner read the
safety/efficiency trade-off as a tangible knob. Rather than tuning a
weighted-sum coefficient until ``it looks reasonable,'' the planner
picks a target survival probability $e^{-\bar{R}}$ and the solver
returns the cheapest fleet plan that meets it (or reports
infeasibility, which is itself useful information). The fleet-wide
residential-edge budget $\bar{H}$ provides a separate dial for
political and community concerns about vulnerable-street penetration,
which is operationally distinct from per-rider safety.

A second operator-facing benefit is calibration auditability. Every
constant in the BASM derivation traces to a public source and a
specific configuration field, so a third party can reproduce the
calibration without contacting the authors and can swap in alternative
priors at the YAML level without code changes.

\section{Limitations}

Three honest limitations remain.

\paragraph{Public-aggregate calibration.}
The BASM uses public aggregate \citet{morth2022} data and an OSM
structural proxy; it does not have access to the protected iRAD
microdata. The per-edge probability $r_{ij}$ is therefore a
calibrated prior, not a measurement. The $\pm 25\%$ city-aggregate
cross-validation of Section~\ref{ch4:xv} bounds drift but does not
turn the prior into ground truth.

\paragraph{Super-graph fixes the path.}
The optimisation is over customer \emph{orderings}, not over which
physical streets to drive on. Once depots and customers are fixed,
the super-arc $(u,v)$ uses the shortest-time path on the underlying
street graph, and $R_{ij}$ and $H_{ij}$ are computed on that fixed
path. A risk-aware planner might, in principle, prefer a slightly
slower path through safer streets; the current formulation cannot
select that. Lifting this restriction is left as future work because
it changes the variable structure of the model fundamentally.

\paragraph{KML cross-validation against OSM.}
The Blinkit KML cross-validation against OpenStreetMap returned an
$8\%$ match rate at 1~km radius. We interpret this as evidence about
sparse OSM brand tagging in India rather than KML inaccuracy, but a
manual sample audit on Google Maps would lift this from a coverage
statistic to an accuracy figure.

A fourth, lesser, limitation is that MILP scaling to $N \geq 100$ is
impractical with CBC; a commercial solver (Gurobi/CPLEX) would close
some of that gap. The heuristics already produce competitive bounds,
and the \texttt{auto} backend in
\texttt{src/savrptw/solvers/milp.py} switches to commercial
solvers when their Python bindings are present.

\section{Threats to Validity}

The customer placement model uses residential-density weighting on
the OSM \texttt{landuse} layer rather than a population-density
raster. Where census or WorldPop microdata is available, plugging it
into \texttt{instance.generator} would tighten the demand model. The
behavioural compliance prior is anchored on a single field-data
citation \citep{das2020}, not a multi-study meta-analysis; the
Beta-prior sensitivity sweep in Chapter~6 is the safeguard against
mis-specification.

\section{Empirical Findings}

The completed experiment grid (428 feasible rows across five cities
and four problem sizes) yields three concrete empirical conclusions.

\textbf{GA dominates ALNS at large $N$.} At $N = 200$, the giant-tour
GA with Bellman--Ford split produces lower mean $F_1$ than ALNS in
every city with more than one seed (e.g.\ Bengaluru: GA $F_1 = 4.89$
vs.\ ALNS $F_1 = 6.94$; Delhi: GA $5.82$ vs.\ ALNS $9.66$), while
running approximately $325\times$ faster (mean 26~s vs.\ 8{,}407~s).
The Bellman--Ford decoder pre-solves the route-assignment sub-problem
at each fitness evaluation, making the GA remarkably effective for
this formulation relative to the destroy-repair operators of ALNS.

\textbf{Both solvers scale near-linearly in $F_1$ with $N$.}
Doubling $N$ from 50 to 100 approximately doubles mean $F_1$
(Bengaluru GA: $1.74 \to 3.35$; ALNS: $1.64 \to 3.98$), consistent
with the soft-time-window penalty accumulating proportionally to the
number of customers served late.

\textbf{City geometry matters.} Gurugram and Mumbai produce elevated
$F_1$ at $N \geq 50$ relative to their Blinkit dark-store counts,
driven by fewer depot clusters relative to demand spread. Bengaluru
provides the most stable per-seed results and is the recommended
primary city for sensitivity studies.

\section{Summary of Contributions}

The contributions of this thesis are as follows.

\begin{enumerate}
    \item A Safety-Aware VRPTW formulation that treats per-route
    crash exposure as a hard $\varepsilon$-constraint rather than as
    an objective component, with a separate fleet-wide residential
    budget ($\bar{H}$) and a per-route residential cap
    ($\bar{H}_k = 100$).
    \item A documented public-source crash-risk calibration chain
    (BASM) anchored on \citet{morth2022} aggregates,
    \citet{mohan2017} ordering constraints, and an OpenStreetMap
    structural proxy; the explicit $\pi_{ij}$ formula in Chapter~4
    means the proxy is reproducible from the manuscript alone.
    \item A unified solver/evaluation contract under which a
    PuLP-based MILP, a giant-tour GA with Bellman--Ford optimal
    split \citep{prins2004}, and an Adaptive Large Neighbourhood
    Search \citep{ropke2006} all evaluate $F_1$ through identical
    code, eliminating the apples-to-pears bias common in VRP
    solver comparisons.
    \item An empirical finding that the GA with Bellman--Ford split
    outperforms ALNS at $N \geq 100$ in both solution quality and
    wall-clock time on the SA-VRPTW formulation, with a $325\times$
    speed advantage at $N = 200$ across five Indian cities.
    \item A five-city empirical study on real Blinkit dark-store
    coordinates from a publicly committed KML snapshot, with
    Hyderabad explicitly excluded because the snapshot has zero
    in-bbox placemarks.
    \item A reproducibility surface that includes Hydra
    configuration, seeded experiment runs, a manifest-checkpointed
    grid driver, and a calibration note that prevents production
    runs from silently using uncalibrated risk constants.
\end{enumerate}

\section{Conclusion}

This thesis presented an SA-VRPTW formulation for q-commerce
last-mile delivery in Indian cities that handles crash safety as an
explicit $\varepsilon$-constraint, calibrates per-edge crash
probability through a documented public-source chain, and is
implemented under a unified solver and evaluation contract across
MILP, GA, and ALNS. The empirical study on 428 feasible instances
across five cities demonstrates that the giant-tour GA with
Bellman--Ford split is both faster and higher-quality than ALNS at
$N \geq 100$, with a $325\times$ wall-clock advantage at $N = 200$.
The full pipeline is publicly released under the MIT licence.

\section{Future Work}

The next phase is operational and empirical rather than conceptual.

\paragraph{Behavioural meta-analysis.}
Replace the single-study \citet{das2020} compliance prior with a
multi-study meta-analysis covering Indian gig-rider populations
beyond Bengaluru.

\paragraph{Real-time congestion.}
Replace the static BPR model with a real-time congestion feed where
available; the BPR model in \texttt{conf/congestion/bpr.yaml} is
already isolated, so the swap is local.

\paragraph{Stochastic demand.}
Extend the formulation to a stochastic demand variant in which order
arrivals across the dispatch window are modelled by a non-stationary
point process rather than a uniform $[0, 5]$~minute draw.

\paragraph{Risk-aware path selection.}
Lift the super-graph restriction so that the optimisation jointly
selects customer orderings and physical street paths. This is a
materially harder model, but it is the natural next step beyond the
current safety-aware-\emph{ordering} formulation.

\paragraph{Direct crash microdata.}
If access to city-level crash microdata or validated black-spot
inventories becomes possible, BASM can be upgraded from a
public-aggregate surrogate to a more direct empirical edge-risk
model without changing the surrounding solver architecture.
