\chapter{Calibration and Data Pipeline}
\label{ch4}

This chapter documents the data pipeline that materialises an
SA-VRPTW instance from public geographic and road-safety data. The
implementation lives under \texttt{src/savrptw/} and is composable
through Hydra, with every constant traceable to a configuration file
under \texttt{conf/}. The pipeline produces an immutable
\texttt{Instance} dataclass that the three solvers in Chapter~5 share
verbatim.

\section{Road Network Extraction}

City street graphs are downloaded from OpenStreetMap using OSMnx
\citep{boeing2017osmnx}. Each city is defined by a committed bounding
box in \texttt{conf/city/<city>.yaml}; the loader extracts the graph
with \texttt{network\_type='drive'} and simplifies it into a directed
multigraph. During enrichment, OSM tags that otherwise appear as
lists are standardised to scalars, missing speeds are imputed via
\texttt{osmnx.add\_edge\_speeds()}, and free-flow travel time is
computed as
\begin{equation}
    t_{ij}^{free} = \frac{60 \cdot \ell_{ij}}{1000 \cdot v_{ij}^{free}}
    \quad \text{(min)}.
\end{equation}
Each edge is also assigned a binary residential indicator $h_{ij}$
that equals 1 for OSM \texttt{highway} tags
\texttt{residential} and \texttt{living\_street}, and 0 otherwise.

\section{BPR Congestion Model}

Travel-time inflation uses the Bureau of Public Roads functional form
\citep{bpr1964}:
\begin{equation}
    \rho_{ij} = 0.15 \left(\frac{V_{ij}}{C_{ij}}\right)^4,
    \quad
    t_{ij} = t_{ij}^{free}\,(1 + \rho_{ij}),
    \label{eq:BPR}
\end{equation}
with capacity $C_{ij}$ from the Highway Capacity Manual
\citep{hcm2010} (Exhibit~10-3) indexed by
\texttt{(highway\_class, lane\_count)} and flow $V_{ij}$ derived from
a class-level annual average daily traffic (AADT) prior multiplied by
an hour-of-day factor (peak windows 08--11 and 18--21). All values
live in \texttt{conf/congestion/bpr.yaml}; reviewers can audit the
calibration without reading source code.

This design has two advantages. First, congestion parameters are
visible to reviewers in configuration rather than buried in code.
Second, all solvers see the same congestion-inflated edge times
through the single field $t_{ij}$, so there is no possibility of
divergent congestion treatment between MILP and heuristics.

\section{BASM: Bengaluru Accident Surrogate Model}
\label{ch4:basm}

The Bengaluru and Indian-city Accident Surrogate Model (BASM)
converts publicly documented aggregate Indian road-safety rates into
a transparent edge-level prior, which is then redistributed within
each OSM road class using graph-local structure. The functional form
is
\begin{equation}
    r_{ij} = \mathrm{clip}\!\left[
        \lambda_{\mathrm{class}(ij)} \cdot \frac{\ell_{ij}}{1000} \cdot
        \sigma_{\mathrm{class}(ij)} \cdot \pi_{ij},
        \,0,\, 0.99\right],
    \label{eq:BASM}
\end{equation}
where:
\begin{itemize}
    \item $\lambda_{\mathrm{class}(ij)}$ is a per-km, per-traversal
    base probability for the OSM road class, derived from
    \citet{morth2022} fatal-plus-grievous events per category-km
    divided by the project's BPR-derived annual traversal exposure;
    \item $\sigma_{\mathrm{class}(ij)} \in [0.5, 2.0]$ is a
    rank-preserving severity multiplier (arterial $>$ collector $>$
    local), computed via the bounded transform
    $\sigma = \sqrt{\text{intensity} / g}$, where intensity is the
    fatality-share-to-length-share ratio per category and $g$ is the
    geometric mean of intensities; this preserves the
    \citet{mohan2017} ordering;
    \item $\pi_{ij} \in [0.5, 2.0]$ is a bounded local proxy
    combining approximate edge betweenness, traffic-signal density,
    and crossing density at the edge endpoints.
\end{itemize}

\paragraph{Explicit local-proxy formula.}
The bounded local proxy $\pi_{ij}$ is computed as
\begin{equation}
    \pi_{ij} = \mathrm{clip}\!\left[
        1 + 0.40\,(2\beta_{ij} - 1) + 0.30\,s_{ij} + 0.30\,c_{ij},
        \,0.5,\, 2.0\right],
    \label{eq:pi-proxy}
\end{equation}
where $\beta_{ij} \in [0,1]$ is the normalised approximate edge
betweenness, $s_{ij} \in [0,1]$ is the local traffic-signal density,
and $c_{ij} \in [0,1]$ is the local crossing density. The weights
$(0.40, 0.30, 0.30)$ are committed to
\texttt{conf/risk/basm\_v1.yaml}; the betweenness weight dominates
because betweenness is a structural correlate of through-traffic
volume, while signal and crossing density share the remaining mass
equally. The clip endpoints $[0.5, 2.0]$ cap how far local structure
can move the prior from the class baseline; this prevents a single
unusual edge from dominating either tail.

\paragraph{What BASM does not do.}
BASM does not have access to the protected Integrated Road Accident
Database (iRAD) microdata. Equation~\eqref{eq:BASM} is therefore a
calibrated prior, not a measurement. The detailed derivation,
including page references for every \citet{morth2022} number used
and the geometric-mean derivation of $\sigma$, is documented in the
project repository at \texttt{docs/BASM\_CALIBRATION.md}.

\section{Cross-Validation Against City-Level Aggregates}
\label{ch4:xv}

The instantiated $r_{ij}$ field for each retained city is
cross-validated against the city's published annual fatal-plus-grievous
totals. For each city we sum
$\sum_{(i,j) \in E} r_{ij} \cdot x_{ij}^{\text{annual}}$ over the
extracted street graph, where $x_{ij}^{\text{annual}}$ is the
BPR-derived annual traversal count for edge $(i,j)$. The resulting
modelled annual events are required to lie within $\pm 25\%$ of the
city's published aggregate. This tolerance reflects the fact that
$\lambda_{\mathrm{class}}$ is a national-level prior while the target
is city-specific; the band absorbs reasonable city-scale variation
without permitting unbounded drift. Cities outside the band are
flagged at instance build time and the calibration is rejected.

\section{Dark Stores}

Depot candidates are derived from the publicly available KML snapshot
at \url{https://darkstoremap.in/dark_store.kml}, committed to the
repository as
\texttt{data/raw/darkstoremap\_in\_2026-04-09.kml}. The committed
snapshot contains 461 Blinkit placemarks; per-city in-bounding-box
counts are summarised in Table~\ref{tab:darkstore-counts}.

\begin{table}[ht]
\centering
\caption{Blinkit dark-store coverage per city in the 2026-04-09 KML
snapshot.}
\label{tab:darkstore-counts}
\begin{tabular}{lc}
\toprule
\textbf{City} & \textbf{In-bbox count} \\
\midrule
Bengaluru & 113 \\
Delhi     & 111 \\
Mumbai    &  96 \\
Pune      &  39 \\
Gurugram  &  38 \\
Hyderabad &   0 (excluded) \\
\bottomrule
\end{tabular}
\end{table}

Hyderabad is excluded from the primary single-brand experiments
because the snapshot has zero in-bbox Blinkit placemarks there;
re-fetching the KML on 2026-04-25 produced a byte-identical file. For
each retained city, k-means clustering with $k \in \{3, 5, 8\}$
depots produces depot centroids, which are then snapped to the
nearest OSM nodes.

\section{OpenStreetMap Cross-Validation}

A secondary validation queries the OSM Overpass API for features
tagged \texttt{brand=Blinkit} or matching \texttt{name~/Blinkit/i}
within radii of 250~m and 1~km of each sampled KML coordinate. At a
1~km radius across 24 stores, only 2 matched ($8\%$, 95\% Wilson
confidence interval $[0.02, 0.26]$ \citep{wilson1927}), driven by a
single hit in Delhi and one in Gurugram. We interpret this result as
evidence about OSM's sparse coverage of Indian dark stores rather
than about KML accuracy, and document the conclusion in the
project repository at \texttt{docs/DARKSTORE\_VALIDATION.md}.

\section{Customer Sampling and Super-Arc Construction}

Customers are sampled from the road graph with residential weighting:
a node adjacent to at least one residential edge receives full
weight, and non-residential nodes receive a smaller background
weight. Each sampled customer is assigned demand $q_i \in \{1, 2\}$,
order release time $e_i \sim \mathcal{U}(0, 5)$~min, promised time
$\mathrm{ETA}_i = e_i + 10$, and a home depot chosen by nearest
free-flow travel time.

Once depots and customers are fixed, a complete super-graph is
built. For each origin super-node, a single-source Dijkstra run on
the congestion-inflated street graph determines the shortest-time
path to every other super-node. The path is then aggregated into the
super-arc tuple $(T_{uv}, R_{uv}, H_{uv})$ via
Equations~\eqref{eq:Tuv}--\eqref{eq:Huv}.

\section{Reproducibility Guard-Rails}

Two guard-rails enforce calibration discipline. First, the BASM
attachment refuses to run if the configuration field
\texttt{source} is marked \texttt{uncalibrated}; this prevents a
production run from silently using arbitrary placeholder constants.
Second, the test suite under \texttt{tests/} verifies that every
attached edge risk satisfies $0 \leq r_{ij} \leq 0.99$ and that the
class ordering arterial $>$ collector $>$ local is preserved on a
controlled synthetic graph. Both checks are wired into the smoke-test
target so the manuscript's calibration claims cannot drift away from
the implementation between revisions.
