\chapter{Problem Formulation}
\label{ch3}

This chapter states the Safety-Aware Vehicle Routing Problem with
Soft Time Windows (SA-VRPTW) in full. The formulation matches the
canonical specification in \texttt{docs/FORMULATION.md} and the
software contract in \texttt{src/savrptw/types.py} on the project
repository. Service quality is optimised directly; route safety and
neighbourhood protection are enforced through explicit
$\varepsilon$-constraints. Crash risk is not a weighted objective
term, congestion is not a separate objective, and there is no
behavioural-fatigue budget inside the optimiser.

\section{Notation}

Let $G = (V, E)$ be the directed street graph for a city, extracted
via OSMnx \citep{boeing2017osmnx} with the drive-network filter. Let
$D \subset V$ be the set of depot nodes (k-means cluster centroids of
the city's Blinkit dark stores, snapped to the nearest OSM nodes),
and $C \subset V$ the set of customer nodes (sampled with
residential-density weighting, constrained to a delivery radius
around depots). Let $K = \bigcup_{d \in D} K_d$ be the rider fleet,
partitioned by home depot $d$.

Table~\ref{tab:notation} summarises every symbol used in the
formulation.

\begin{table}[ht]
\centering
\caption{Notation used in the SA-VRPTW formulation.}
\label{tab:notation}
\small
\begin{tabular}{lll}
\toprule
\textbf{Symbol} & \textbf{Meaning} & \textbf{Source} \\
\midrule
$\ell_{ij}$         & street-edge length (m) & OSMnx \\
$v_{ij}^{free}$     & free-flow speed (km/h) & OSMnx imputed \\
$t_{ij}^{free}$     & free-flow travel time (min) & derived \\
$\rho_{ij}$         & BPR congestion factor & \texttt{conf/congestion/bpr.yaml} \\
$t_{ij}$            & congestion-inflated travel time (min) & derived \\
$r_{ij}$            & per-traversal crash probability & BASM (Ch.~4) \\
$h_{ij}$            & 1 if residential edge, else 0 & OSM \texttt{highway} tag \\
$T_{uv}$            & cumulative travel time on super-arc & sum on shortest path \\
$R_{uv}$            & cumulative log-survival on super-arc & $\sum -\ln(1-r_e)$ \\
$H_{uv}$            & residential-edge count on super-arc & sum on shortest path \\
$q_i$               & demand at customer $i \in \{1,2\}$ & sampled \\
$e_i$               & order release time (min) & sampled $\sim \mathcal{U}(0,5)$ \\
$\mathrm{ETA}_i$    & promised arrival $= e_i + 10$ (min) & ten-minute promise \\
$s_i$               & service time at customer $i \in \{2,4\}$ (min) & high-rise flag \\
$Q$                 & rider capacity (items) & $=2$ \\
$T_{\max}$          & maximum route duration (min) & $=35$ \\
$\bar{R}$           & per-route survival budget & swept (Ch.~6) \\
$\bar{H}_k$         & per-route residential-edge cap & $=100$ \\
$\bar{H}$           & fleet-wide residential-edge budget & swept (Ch.~6) \\
$\beta_{\text{stw}}$ & SLA-penalty curvature & $=0.12$ \\
$w_{\text{early}}$  & early-arrival idle weight & $=1.0$ \\
$M$                 & big-M constant & $\geq T_{\max}$ \\
\bottomrule
\end{tabular}
\end{table}

For each street-graph edge $(i,j) \in E$, the model stores
$\ell_{ij}$, $v_{ij}^{free}$,
$t_{ij}^{free} = 60\ell_{ij}/(1000 v_{ij}^{free})$, the BPR
congestion factor $\rho_{ij}$, the congestion-inflated travel time
$t_{ij} = t_{ij}^{free}(1 + \rho_{ij})$, the per-traversal crash
probability $r_{ij} \in [0, 0.99]$ (Section~\ref{ch4:basm}), and the
residential indicator $h_{ij} \in \{0,1\}$.

For each pair $(u, v)$ with $u, v \in C \cup D$, the super-graph arc
$(u, v)$ is the shortest-time path on $G$. The cumulative quantities
along that path are
\begin{align}
T_{uv} &= \sum_{e \in P_{uv}} t_e, \label{eq:Tuv} \\
R_{uv} &= \sum_{e \in P_{uv}} -\ln(1-r_e), \label{eq:Ruv} \\
H_{uv} &= \sum_{e \in P_{uv}} h_e. \label{eq:Huv}
\end{align}
$R_{uv}$ is additive in log-survival because per-traversal crashes
are treated as independent events along the path. The super-graph
fixes the underlying street-graph path for each super-arc at instance
generation time, which means the optimisation is over customer
\emph{orderings}, not over which physical streets to drive on. This
restriction is intentional and is discussed in the limitations
(Chapter~7).

\section{Customer and Fleet Parameters}

Each customer $i \in C$ has demand $q_i \in \{1,2\}$, an order
release time $e_i \in [0, 5]$~min, a promised arrival
$\mathrm{ETA}_i = e_i + 10$ (the q-commerce ten-minute promise), and
a service time $s_i \in \{2, 4\}$~min (4~min for high-rise buildings,
otherwise 2). Time windows are soft. The rider fleet is homogeneous
with capacity $Q = 2$ items and maximum route duration
$T_{\max} = 35$~min. Per-depot rider counts are sized as
$|K_d| = \lceil 1.2 N_d / Q \rceil$, providing a 20\% over-provision
buffer above the no-slack lower bound.

\section{Decision Variables}

For rider $k \in K$ and super-arc $(i, j)$:
\begin{itemize}
    \item $x_{ij}^k \in \{0, 1\}$: rider $k$ traverses super-arc $(i, j)$;
    \item $a_i^k \geq 0$: arrival time of rider $k$ at node $i$ (defined only when $i$ is visited);
    \item $w_i^k \geq 0$: idle wait at $i$ (early arrival);
    \item $\tau_i^k \geq 0$: lateness at $i$;
    \item $u_i^k \in \mathbb{Z}_{\geq 1}$: Miller--Tucker--Zemlin
    sequence variable for sub-tour elimination \citep{miller1960}.
\end{itemize}

\section{Primary Objective}

The primary objective $F_1$ minimises early-arrival waiting plus the
convex SLA lateness penalty across the fleet:
\begin{equation}
    F_1 = \sum_{k \in K} \sum_{i \in C}
    \left[ w_{\text{early}} \cdot w_i^k +
    \big(\exp(\beta_{\text{stw}} \tau_i^k) - 1\big) \right].
    \label{eq:F1}
\end{equation}
Travel time enters implicitly through the arrival-time linking
constraints; congestion is already baked into $t_{ij}$; crash risk is
handled as an $\varepsilon$-constraint, not as an objective term, so
the operator can read its cost off the Pareto frontier.

\section{$\varepsilon$-Constraints}
\label{ch3:eps}

Per-route crash-survival budget:
\begin{equation}
    \sum_{(i,j)} R_{ij} \, x_{ij}^k \leq \bar{R}, \quad \forall k \in K.
    \label{eq:Rbar}
\end{equation}
Equation~\eqref{eq:Rbar} has a clear operational interpretation: the
route survival probability must be at least $\exp(-\bar{R})$.

Fleet-wide residential-edge budget:
\begin{equation}
    \sum_{k \in K} \sum_{(i,j)} H_{ij} \, x_{ij}^k \leq \bar{H}.
    \label{eq:Hbar}
\end{equation}

Per-route residential cap:
\begin{equation}
    \sum_{(i,j)} H_{ij} \, x_{ij}^k \leq \bar{H}_k, \quad \forall k \in K.
    \label{eq:Hk}
\end{equation}
The implementation uses $\bar{H}_k = 100$ residential-edge
traversals per route. The pair $(\bar{R}, \bar{H})$ is swept on a
finite grid (Chapter~6) to trace the Pareto frontier; non-dominated
points are retained.

\section{Core VRPTW Constraints}

Each customer is visited exactly once:
\begin{equation}
    \sum_{k \in K} \sum_{j \in C \cup D,\, j \neq i} x_{ji}^k = 1,
    \quad \forall i \in C. \label{eq:visit}
\end{equation}
Flow conservation at every customer:
\begin{equation}
    \sum_{i} x_{ij}^k = \sum_{i} x_{ji}^k, \quad \forall j \in C, \forall k \in K. \label{eq:flow}
\end{equation}
Capacity:
\begin{equation}
    \sum_{i \in C} q_i \sum_{j} x_{ij}^k \leq Q, \quad \forall k \in K. \label{eq:cap}
\end{equation}

\section{Multi-Depot Coupling}

Riders are partitioned by home depot. A rider $k \in K_d$ assigned to
depot $d$ leaves $d$ at most once, returns iff it left, and cannot
touch any other depot. Formally, for every $k \in K_d$:
\begin{align}
    \sum_{j \in C} x_{dj}^k &\leq 1, \label{eq:depot-leave} \\
    \sum_{j \in C} x_{jd}^k &= \sum_{j \in C} x_{dj}^k, \label{eq:depot-return} \\
    x_{uj}^k = 0 \quad &\forall u \in D \setminus \{d\}, \forall j, \label{eq:no-other-depot-out} \\
    x_{ju}^k = 0 \quad &\forall u \in D \setminus \{d\}, \forall j. \label{eq:no-other-depot-in}
\end{align}
Equation~\eqref{eq:depot-leave} bounds the dispatch count to one;
\eqref{eq:depot-return} couples dispatch and return; and the two
constraints~\eqref{eq:no-other-depot-out}--\eqref{eq:no-other-depot-in}
forbid touching any depot other than the home depot.

\section{Sub-Tour Elimination and Arrival Linking}

Sub-tour elimination follows the Miller--Tucker--Zemlin construction
\citep{miller1960}, with coefficient $|C|$ rather than vehicle
capacity:
\begin{align}
    u_i^k - u_j^k + |C| \, x_{ij}^k &\leq |C| - 1,
    \quad \forall i, j \in C, i \neq j, \forall k, \label{eq:mtz} \\
    1 \leq u_i^k &\leq |C|, \quad \forall i \in C, \forall k.
\end{align}

Arrival times are linked by big-$M$ constraints:
\begin{equation}
    a_j^k \geq a_i^k + s_i + T_{ij} - M(1 - x_{ij}^k),
    \quad \forall i, j, k. \label{eq:link}
\end{equation}
The big-$M$ is set to a tight value derived from $T_{\max}$ rather
than a hard-coded 240; this materially reduces the LP relaxation
gap for CBC and prevents the spurious infeasibility observed in
earlier drafts of the implementation.

The maximum route-duration constraint is
\begin{equation}
    \sum_{(i,j)} T_{ij} \, x_{ij}^k +
    \sum_{i} s_i \sum_{j} x_{ji}^k \leq T_{\max}, \quad \forall k. \label{eq:Tmax}
\end{equation}

Soft-window slack variables are linked via
\begin{align}
    w_i^k &\geq e_i - a_i^k, \quad w_i^k \geq 0, \label{eq:early} \\
    \tau_i^k &\geq a_i^k - \mathrm{ETA}_i, \quad \tau_i^k \geq 0. \label{eq:late}
\end{align}

\section{MILP Linearisation of the SLA Penalty}
\label{ch3:pwl}

The mixed-integer linear program cannot evaluate
$\exp(\beta_{\text{stw}}\tau)$ directly. The convex term
$\exp(\beta_{\text{stw}}\tau) - 1$ is approximated via a
piecewise-linear outer envelope at breakpoints
\[
    \tau \in \{0, 2, 5, 10, 15, 20, 30\} \text{ minutes},
\]
encoded with SOS2 sets so that at most two adjacent breakpoints are
active per constraint. The breakpoints were chosen by a
curvature-density heuristic to keep the relative approximation error
below 5\% over $\tau \in [0, 30]$ for the calibrated curvature
$\beta_{\text{stw}} = 0.12$. The heuristic concentrates breakpoints
where the second derivative is largest (small $\tau$) and spaces them
out in the long-tail region.

The heuristics evaluate the exponential exactly. Both the MILP and
the heuristics share the same feasibility checker
(\texttt{savrptw.eval.feasibility}) and report $F_1$ through the same
\texttt{savrptw.eval.objective} module.

\section{Implementation Constants}

The default constants in
\texttt{conf/config.yaml} and \texttt{conf/instance/default.yaml} are:
$Q = 2$ items;
$T_{\max} = 35$~min;
$\bar{H}_k = 100$ residential-edge traversals per route;
$\beta_{\text{stw}} = 0.12$;
$w_{\text{early}} = 1.0$;
$M$ tight to $T_{\max}$ plus the longest super-arc.
The fleet-wide residential budget $\bar{H}$ defaults to $10^6$
(an effective $+\infty$ sentinel) and is tightened in the
$\varepsilon$-sweep of Chapter~6. The per-route survival budget
$\bar{R}$ similarly defaults to $+\infty$ in the unconstrained
benchmark and is swept across $\{0.05, 0.10, 0.20, 0.40, 0.80, 1.60,
\infty\}$ for the Pareto study.
