\chapter{Conclusion and Future Work}

\section{Summary of Contributions}

This thesis update consolidates the corrected SA-VRPTW formulation and the corresponding software architecture. The key contributions of the current branch are as follows.

\begin{enumerate}
    \item \textbf{Corrected objective structure.} The model now optimizes service quality through waiting and exponential lateness penalties, while route risk and residential exposure are enforced through explicit epsilon-constraints rather than through a weighted sum of incompatible units.
    \item \textbf{Reproducible BASM calibration.} The earlier placeholder risk layer has been replaced with a documented calibration based on MoRTH 2022 aggregates, Mohan et al.'s arterial-versus-non-arterial directionality, and graph-local OpenStreetMap proxies.
    \item \textbf{Consistent solver contract.} The exact model, genetic algorithm, and adaptive large neighbourhood search are all tied to the same instance representation, feasibility checks, and post-hoc evaluators.
    \item \textbf{Transparent experiment scope.} The primary single-brand city set has been fixed to Bengaluru, Delhi, Gurugram, Mumbai, and Pune, with Hyderabad excluded explicitly because the committed Blinkit snapshot does not cover it.
    \item \textbf{Thesis and code alignment.} The manuscript no longer claims DRL baselines, six-city primary experiments, synthetic iRAD calibration, or unmeasured figure results that are absent from the current branch.
\end{enumerate}

\section{Current Findings}

The strongest conclusion supported by the current branch is methodological rather than final empirical. A safety-aware q-commerce routing problem can be expressed cleanly without conflating travel time, congestion, and crash risk into one arbitrary weighted objective. The branch demonstrates that a public-source calibration path for edge-level crash probabilities is implementable, that the solver stack can consume a shared canonical instance, and that the experimental protocol can be defined without relying on unverifiable baselines.

\section{Limitations}

Several limitations of the current work are acknowledged:

\begin{itemize}
    \item The full five-city experiment grid has not yet been executed on the corrected formulation, so this manuscript update avoids claiming final numerical trade-offs.
    \item The BASM calibration is publicly reproducible, but it still depends on aggregate road-safety data rather than direct city-level crash microdata for every retained city.
    \item The local OSM proxy redistributes risk within road classes using graph structure rather than observed crash black spots, which is defensible but still indirect.
    \item Behavioural compliance is currently evaluated post hoc through sensitivity analysis rather than estimated from platform-specific telematics or rider logs.
\end{itemize}

\section{Future Research Directions}

The next phase of work is operational and empirical rather than conceptual.

\vspace{0.5em}
\noindent\textbf{Five-city target validation.}
The immediate requirement is to complete the city-level validation dataset so that the calibrated BASM graph for each retained city can be checked against a public annual target within the accepted tolerance band.

\vspace{0.5em}
\noindent\textbf{Full experiment execution and figure regeneration.}
The epsilon-grid must be executed across all retained cities, problem sizes, and seeds, after which the manuscript figures can be generated directly from the structured JSON outputs and paired CSV artefacts.

\vspace{0.5em}
\noindent\textbf{Richer behavioural evidence.}
The Monte Carlo compliance model should eventually be informed by rider telemetry, platform audit data, or structured survey evidence rather than by sensitivity sweeps alone.

\vspace{0.5em}
\noindent\textbf{Direct crash microdata integration.}
If access to city-level crash microdata or validated black-spot inventories becomes possible, BASM can be upgraded from a public-aggregate surrogate to a more direct empirical edge-risk model without changing the surrounding solver architecture.
