\thispagestyle{plain}
\begin{center}
    \Large \textbf{\uppercase{Abstract}}
\end{center}

\vspace{2\baselineskip}

\noindent
Quick-commerce platforms in India promise ten-minute delivery from
neighbourhood ``dark stores'' to customers, putting persistent time
pressure on a fleet of two-wheeler riders that must operate within an
already hazardous urban traffic environment. Conventional vehicle
routing formulations minimise travel time and treat safety as a
post-hoc concern. This thesis proposes a Safety-Aware Vehicle Routing
Problem with Soft Time Windows (SA-VRPTW) that explicitly bounds
per-route crash exposure and fleet-wide use of vulnerable residential
roads via $\varepsilon$-constraints, while minimising a convex
soft-time-window penalty as the primary objective.

The crash risk per road segment is calibrated through a documented
public-source chain---the Ministry of Road Transport and Highways
\citep{morth2022} aggregate fatality counts, the
\citet{mohan2017} arterial-versus-non-arterial directionality, and a
bounded OpenStreetMap structural proxy combining edge betweenness,
traffic-signal density, and crossing density---without requiring
access to the protected Integrated Road Accident Database microdata.
Travel times are inflated by a Bureau of Public Roads
\citep{bpr1964} congestion model with class-level capacities indexed
into the Highway Capacity Manual \citep{hcm2010}.

Three solvers are implemented behind a unified evaluation contract:
a piecewise-linear MILP, a giant-tour genetic algorithm with
Bellman--Ford optimal split decoding \citep{prins2004}, and an
Adaptive Large Neighbourhood Search with Ropke--Pisinger adaptive
operator weights \citep{ropke2006}. All three consume a typed
\texttt{Instance} dataclass and report $F_1$ through identical
evaluation code, eliminating the apples-to-pears bias common in VRP
solver comparisons.

The empirical study benchmarks five Indian cities (Bengaluru, Delhi,
Gurugram, Mumbai, Pune) with real Blinkit dark-store coordinates
extracted from a publicly committed KML snapshot, sweeping
$N \in \{20, 50, 100, 200\}$ over multiple seeds. The protocol traces
Pareto frontiers between $F_1$ delivery performance and per-route
crash survival, quantifies the cost of the residential-road
constraint, and verifies behavioural sensitivity through a Beta-prior
compliance Monte Carlo. The contribution is a fully reproducible,
calibration-honest framework that lets a quick-commerce operator dial
the safety/efficiency trade-off explicitly rather than implicitly.

\vspace{\baselineskip}

\noindent
\textbf{Keywords}: Safety-aware vehicle routing, soft time windows,
$\varepsilon$-constraint multi-objective optimisation,
quick-commerce, last-mile delivery, mixed-integer linear programming,
genetic algorithms, adaptive large neighbourhood search, crash-risk
calibration, OpenStreetMap.
