\thispagestyle{plain}

\begin{center}
 \Large {\bf \uppercase{Acknowledgements}}
\end{center}

\vspace{3\baselineskip}

\noindent
I would like to express my sincere gratitude to my thesis supervisor, \textbf{Prof.\ Subhajit Sidhanta}, Department of Industrial and Systems Engineering, IIT Kharagpur, for his invaluable guidance, constructive feedback, and constant encouragement throughout the course of this research. His insights on optimization methodology and systems thinking were instrumental in shaping the direction of this work.

\vspace{\baselineskip}

\noindent
I am grateful to the faculty and staff of the Department of Industrial and Systems Engineering for providing an excellent academic environment and the computational resources necessary for this research.

\vspace{\baselineskip}

\noindent
I also acknowledge the open-source communities behind OpenStreetMap, OSMnx, NetworkX, PuLP, and PyTorch, whose tools formed the backbone of the computational pipeline developed in this thesis.

\vspace{\baselineskip}

\noindent
Finally, and most deeply, I thank my family for their unconditional love and unwavering support throughout this journey. My father, \textbf{Dr.\ Pradip Bhadane}, has been a constant source of inspiration and strength. My mother, \textbf{Swati Bhadane}, whose patience and encouragement sustained me through every difficult moment. My sister, \textbf{Swara Bhadane}, for her cheerfulness and support that kept me grounded. This work would not have been possible without them.

\noindent
\vspace{\baselineskip} \\
\textbf{\myname} \\
IIT Kharagpur \\
\mydegreedate
